---
title: "Example analysis"
format: html
---

# Background

I will be doing a short data analysis on the built in air quality dataset in R. This dataset contains air quality measurements in New York between May and September in 1973. It contains 6 variables and 153 observations

```{r}
#| label: "setup"
#| include: false
library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r}
#| echo: true
data(airquality)
dim(airquality)
head(airquality, 3)
```

Our question is: does temperature affect the ozone level?

The intended audience for this example analysis is anyone who wants to get a taste of the exploratory data analysis.

Data dictionary: https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality

Let's first play with the dataset and get a taste of how it looks like using `dyplr` package.

To simplify the analysis, I will exclude the missing values.

```{r}
air <- airquality[rowSums(is.na(airquality)) == 0,]
```

After removing the missing values, we have 111 observations and 6 variables

```{r}
dim(air)
```

```{r}
air %>% summarise(mean_ozone = mean(Ozone),
                  median_ozone = median(Ozone),
                  sd_ozone = sd(Ozone),
                  max_ozone = max(Ozone),
                  min_ozone = min(Ozone))
```

Our ozone level is a very skewed dataset. Let's take a look at the histogram

```{r}
ggplot(air, aes(x=Ozone)) + geom_histogram() + 
  ggtitle("Histogram of Ozone Level") + 
  geom_vline(aes(xintercept=mean(Ozone)),
            color="blue", linetype="dashed", size=1) + 
  geom_vline(aes(xintercept=median(Ozone)),
            color="red", linetype="dashed", size=1)
  
```

The blue dashed line is the mean while the red dashed line is the median of ozone level. Our dataset is very skewed.

Mean ozone from May to September is 42, median at 31, maximum at 168, and minimum at 1

```{r}
air  %>%
  select(Ozone, Month, Day)%>% 
  slice_max(Ozone, n = 3)
```

The highest Ozone occurs in August and July.

```{r}
air  %>%
  select(Ozone, Month, Day)%>% 
  slice_min(Ozone, n = 3)
```

The lowest Ozone level occurs in May

```{r}
air %>% filter(Ozone > 100)
```

There are 7 days with Ozone level higher than 100, 6 out of 7 occurs in July and August with temperature over 80 degrees.

It looks like time in the year might affect Ozone level. So let's check it out.

```{r}
air %>% 
  group_by(Month) %>%
  summarise(mean_ozone = mean(Ozone),
            mean_temp = mean(Temp),
            sd_temp = sd(Temp),
            median_temp = median(Temp),
            max_temp = max(Temp),
            min_temp = min(Temp),
            count = n()) %>%
  arrange(desc(mean_temp))
```

July has the highest mean, median, and standard deviation of temperature, but August is only a little bit behind in mean tempature. August has slightly higher mean Ozone level compared to July. May has the lowest mean Ozone and mean temperature

::: callout-tip
In scatterplots, flexibly play with color, dot size, dot shapes to add more information to the plots
:::

```{r}
ggplot(air, aes(x=Temp, y=Ozone)) + 
  geom_point(aes(col=Month)) + ggtitle("Ozone level against temperature")
```

```{r}
ggplot(air, aes(x=Solar.R, y=Ozone)) + 
  geom_point(aes(col=Month)) + ggtitle("Ozone level against solar radiation")
```

```{r}
ggplot(air, aes(x=Wind, y=Ozone)) + 
  geom_point(aes(col=Month))  + ggtitle("Ozone level against average wind speed")
```

```{r}
#| label: fig-ozone-days
#| fig-cap: "We plot ozone levels across days separated by months. There is no strong relationship between the ozone levels and the specific days in the month."
ggplot(air, aes(x=Day, y=Ozone)) + 
  geom_point()  + facet_grid(cols = vars(Month))  + ggtitle("Ozone level across days")
```

::: callout-note
We have used `select`, `slice_max`, `group_by`, `summarize`, `filter`, `arrange` in the `dplyr` package
:::
